---
title: "Comparing detection efficiencies across experiments"
output: rmarkdown::html_vignette
author: "Amy D Willis and David S Clausen"
vignette: >
  %\VignetteIndexEntry{compare-experiments}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Scope and purpose

In this vignette, we will walk through how to use `tinyvamp` to estimate and compare detection efficiencies across batches or experiments. We will consider the Phase 2 data of Costea et al. (2017), who spiked-in a synthetic community to 10 samples, which were then sequenced using three shotgun metagenomic sequencing experimental protocols (protocols H, Q and W). Taxon abundances were also estimated using flow cytometry. 

Our analysis here will estimate the detection efficiencies of the three shotgun protocols compared to flow cytometry. We will also test the null hypothesis that detection efficiencies of the three shotgun protocols are the same using a likelihood ratio test. 

Because of its flexibility, fitting a model using `tinyvamp` involves specifying a lot of parameters. The focus of this vignette will therefore be on *how* we set up the model to estimate detection efficiencies, rather than *why* we set up the model in this way - please see the preprint and references therein for those details!

# Comparing detection efficiencies across experiments

## Setup

We will start by loading the relevant packages and data, and inspecting the data. This dataset contains the MetaPhlAn2-estimated relative abundances of taxa in 29 samples, formatted in the usual MetaPhlAn2 way (percentages listed at all taxonom): 

```{r setup}
library(tidyverse)
library(magrittr)
library(tinyvamp)
data("costea2017_metaphlan2_profiles")
costea2017_metaphlan2_profiles
```

Since we are going to compare relative detections of the taxa in the synthetic community that was spiked into all samples, we will start by filtering to only the species in the spiked-in community (10 taxa): 


```{r}
costea2017_metaphlan2_profiles_species <- costea2017_metaphlan2_profiles %>%
  filter(sapply(Clade,function(x) grepl("s__",x,fixed = TRUE))) %>%
  filter(sapply(Clade, function(x) !grepl("t__",x,fixed = TRUE))) %>%
  filter(sapply(Clade,function(x)
    grepl("saccharolyticum",x,fixed = TRUE) |
      grepl("perfringens",x,fixed = TRUE) |
      grepl("melaninogenica",x,fixed = TRUE)|
      grepl("difficile",x,fixed = TRUE)|
      grepl("enterica",x,fixed = TRUE)|
      grepl("plantarum",x,fixed = TRUE)|
      grepl("cholerae",x,fixed = TRUE)|
      grepl("pseudotuberculosis",x,fixed = TRUE)|
      grepl("Blautia_hansenii",x,fixed = TRUE)|
      grepl("nucleatum",x,fixed = TRUE)
  ))
```

We can now see our final relative abundance table as follows: 

```{r}
costea2017_metaphlan2_profiles_species
```

We now need to connect the 

```{r}
load("/Users/adwillis/research/tinyvamp/costea2017_mock_composition.rda")
load("costea2017_mock_composition.rda")
mock_cols <- costea2017_mock_composition$Taxon

mock_mat <- t(as.matrix(costea2017_mock_composition[,-1]))
colnames(mock_mat) <- mock_cols
```

# Construct observation matrix W

```{r}
cell_concentration <- costea2017_mock_composition

W <- costea2017_mock_composition %>%
  as.data.frame() %>%
  (function(x){ colnames(x)[2] <- "cells_per_ml"
  return(x)}) %>%
  group_by(Taxon) %>%
  summarize(cells_per_ml = mean(cells_per_ml)) %>%
  (function(x){ m <- matrix(x$cells_per_ml,nrow = 1)
  colnames(m) <- x$Taxon
  return(m)})

costea2017_mock_composition %>% 
  rename(cells_per_ml = 2) %>%
  group_by(Taxon) %>%
  summarize(cells_per_ml = mean(cells_per_ml)) %>%
  pull(cells_per_ml)


W_reorder <- sapply(colnames(W),
                    function(x) which(sapply(costea2017_metaphlan2_profiles_species$Clade,
                                             function(d) grepl(x,d,fixed = TRUE)))) %>%
  as.numeric()

W <- rbind(W,
           costea2017_metaphlan2_profiles_species[W_reorder,-1] %>%
             as.matrix() %>%
             t)

``` 

# Specify Designs, etc. to fit full and null models 

```{r} 

protocols <- sapply(rownames(W)[-1],
                    function(x) costea2017_sample_data$Protocol[
                      costea2017_sample_data$Run_accession == x
                    ])

individuals <- sapply(rownames(W)[-1],
                      function(x) costea2017_sample_data$Individual[
                        costea2017_sample_data$Run_accession == x
                      ])

X <- lapply(protocols,
            function(x) as.numeric(
              c(x == "H",x == "Q", x == "W")
            ))
X <- do.call(rbind,X)
X[,1] <- 1

X <- rbind(0,X)

Z <- matrix(1,nrow = 30, ncol = 1)

Z_tilde <- matrix(0,nrow= 30, ncol = 1)
Z_tilde_gamma_cols <- 1

gammas <- apply(W,1,function(x) log(sum(x)))
gammas_fixed_indices <- rep(FALSE, length(gammas))

P <- matrix(W[1,]/sum(W[1,]),nrow =1, ncol = 10)

P_fixed_indices <- matrix(FALSE,nrow = 1, ncol = 10)

X_tilde <- matrix(0,ncol = 3, nrow= 1)

B <- matrix(0,
            ncol = 10,
            nrow = 3)

B_fixed_indices <- matrix(FALSE,ncol = 10,nrow = 3)
B_fixed_indices[,10] <- TRUE


P_tilde <- P*0
P_tilde_fixed_indices <- !P_fixed_indices
gamma_tilde <- matrix(0,ncol = 1, nrow = 1)
gamma_tilde_fixed_indices <- TRUE

``` 

Fit full and null models 

```{r} 
full_model <- estimate_parameters(W = W,
                                  X = X,
                                  Z = Z,
                                  Z_tilde = Z_tilde,
                                  Z_tilde_gamma_cols = 1,
                                  gammas = gammas,
                                  gammas_fixed_indices = gammas_fixed_indices,
                                  P = P,
                                  P_fixed_indices = P_fixed_indices,
                                  B = B,
                                  B_fixed_indices = B_fixed_indices,
                                  X_tilde = X_tilde,
                                  P_tilde = P_tilde,
                                  P_tilde_fixed_indices = P_tilde_fixed_indices,
                                  gamma_tilde = gamma_tilde,
                                  gamma_tilde_fixed_indices = gamma_tilde_fixed_indices,
                                  alpha_tilde = NULL,
                                  Z_tilde_list = NULL,
                                  barrier_t = 1, #starting value of reciprocal barrier penalty coef.
                                  barrier_scale = 10, #increments for value of barrier penalty
                                  max_barrier = 1e12, #maximum value of barrier_t
                                  initial_conv_tol = 1000,
                                  final_conv_tol = 0.1,
                                  constraint_tolerance = 1e-10,
                                  hessian_regularization = 0.01,
                                  criterion = "Poisson",
                                  profile_P = FALSE,
                                  profiling_maxit = 25,
                                  wts = NULL,
                                  verbose = FALSE)

### Create null model specification for test
null_model <- full_model

### set second and third rows of B equal to zero
null_model$B[2:3,] <- 0
null_model$B_fixed_indices[2:3,] <- TRUE
```

This next step can take ages! Good to parallelize if possible. 

```{r}
bootstrap_test <-  bootstrap_lrt(W = W,
                                 fitted_model= full_model,
                                 null_param = null_model,
                                 n_boot = 1000,
                                 ncores = 5, 
                                 parallelize = TRUE)

### fit reparametrization with X_i = [1_H 1_Q 1_W]
### to (more easily) get CIs for protocol-specific effects
### (also can bootstrap from original model, but currently a bit more
### involved to pull out cis for quantities of form A%*%beta than for
### beta itself)
X_repar <- X
X_repar[,1] <- X_repar[,1] - X_repar[,2] - X_repar[,3]
full_reparam  <-
  estimate_parameters(W = W,
                      X = X_repar,
                      Z = Z,
                      Z_tilde = Z_tilde,
                      Z_tilde_gamma_cols = 1,
                      gammas = gammas,
                      gammas_fixed_indices = gammas_fixed_indices,
                      P = P,
                      P_fixed_indices = P_fixed_indices,
                      B = B,
                      B_fixed_indices = B_fixed_indices,
                      X_tilde = X_tilde,
                      P_tilde = P_tilde,
                      P_tilde_fixed_indices = P_tilde_fixed_indices,
                      gamma_tilde = gamma_tilde,
                      gamma_tilde_fixed_indices = gamma_tilde_fixed_indices,
                      alpha_tilde = NULL,
                      Z_tilde_list = NULL,
                      barrier_t = 1, #starting value of reciprocal barrier penalty coef.
                      barrier_scale = 10, #increments for value of barrier penalty
                      max_barrier = 1e12, #maximum value of barrier_t
                      initial_conv_tol = 1000,
                      final_conv_tol = 0.1,
                      constraint_tolerance = 1e-10,
                      hessian_regularization = 0.01,
                      criterion = "Poisson",
                      profile_P = FALSE,
                      profiling_maxit = 25,
                      wts = NULL,
                      verbose = TRUE)

full_cis <- bootstrap_ci(W,
                         fitted_model = full_reparam,
                         n_boot=1000,
                         m = NULL,
                         alpha = 0.05,
                         parallelize = TRUE,
                         ncores = 5,
                         seed = 3423,
                         return_models = FALSE,
                         verbose = FALSE
                         
)

taxa <- colnames(W)
taxa <- lapply(taxa,
               function(x) strsplit(x,"_") %>%
                 (function(y) paste(substr(y[[1]][1],1,1),". ",
                                    y[[1]][2],sep = "",collapse = ""))) %>%
  do.call(c,.)

full_cis$ci %>%
  filter(param == "B") %>%
  mutate(Protocol = c("H","Q","W")[k],
         Estimate = round(value,2)) %>%
  mutate(Upper = round(upper_ci,2),
         Lower = round(lower_ci,2),
         Taxon = rep(taxa[1:9],3)) %>%
  dplyr::select(Protocol, Taxon, Estimate, Lower, Upper) %>%
  mutate(Estimate = apply(cbind(Estimate,Lower,Upper),
                          1, function(x) paste(x[1]," (", x[2], " - ", x[3], ")",
                                               sep = "",
                                               collapse = ""))) %>%
  select(c(Protocol, Taxon, Estimate)) %>%
  pivot_wider(id_cols = Taxon, names_from = Protocol,
              values_from = Estimate) %>%
  knitr::kable(format = "latex")


``` 

# 10-fold cross-validation 

```{r} 

# construct folds
folds <- vector(10, mode = "list")
available <- 2:30
unique_individuals <- c(1:8,"M")

for(i in 1:9){
  folds[[i]] <- which(individuals == unique_individuals[i]) +1
}

folds[[10]] <- which(individuals %in% c("A","B")) +1

full_cv <- vector(10, mode = "list")
null_cv <- vector(10, mode = "list")
for(whichfoldout in 1:10){
  print(whichfoldout)
  heldout <- folds[[whichfoldout]]
  nheldout <- length(heldout)
  Z_cv <- Z
  Z_cv <- cbind(Z_cv - Z_cv*(1:30 %in% heldout))
  P_cv <- P
  P_fixed_indices_cv <- P_fixed_indices
  for(k in 1:nheldout){
    Z_cv <- cbind(Z_cv,as.numeric(1:30 == heldout[k]))
    P_cv <- rbind(P_cv,P)
    P_fixed_indices_cv <- rbind(P_fixed_indices_cv,
                                P_fixed_indices)
  }
  
  full_cv[[whichfoldout]]  <-
    estimate_parameters(W = W,
                        X = X,
                        Z = Z_cv,
                        Z_tilde = Z_tilde,
                        Z_tilde_gamma_cols = 1,
                        gammas = gammas,
                        gammas_fixed_indices = gammas_fixed_indices,
                        P = P_cv,
                        P_fixed_indices = P_fixed_indices_cv,
                        B = B,
                        B_fixed_indices = B_fixed_indices,
                        X_tilde = X_tilde,
                        P_tilde = P_tilde,
                        P_tilde_fixed_indices = P_tilde_fixed_indices,
                        gamma_tilde = gamma_tilde,
                        gamma_tilde_fixed_indices = gamma_tilde_fixed_indices,
                        alpha_tilde = NULL,
                        Z_tilde_list = NULL,
                        barrier_t = 1, #starting value of reciprocal barrier penalty coef.
                        barrier_scale = 10, #increments for value of barrier penalty
                        max_barrier = 1e12, #maximum value of barrier_t
                        initial_conv_tol = 1000,
                        final_conv_tol = 0.1,
                        constraint_tolerance = 1e-10,
                        hessian_regularization = 0.01,
                        criterion = "Poisson",
                        profile_P = FALSE,
                        profiling_maxit = 25,
                        wts = NULL,
                        verbose = FALSE)
  
  null_cv[[whichfoldout]]  <-
    estimate_parameters(W = W,
                        X = X[,1,drop = FALSE],
                        Z = Z_cv,
                        Z_tilde = Z_tilde,
                        Z_tilde_gamma_cols = 1,
                        gammas = gammas,
                        gammas_fixed_indices = gammas_fixed_indices,
                        P = P_cv,
                        P_fixed_indices = P_fixed_indices_cv,
                        B = B[1,,drop = FALSE],
                        B_fixed_indices = B_fixed_indices[1,,drop = FALSE],
                        X_tilde = X_tilde[,1,drop = FALSE],
                        P_tilde = P_tilde,
                        P_tilde_fixed_indices = P_tilde_fixed_indices,
                        gamma_tilde = gamma_tilde,
                        gamma_tilde_fixed_indices = gamma_tilde_fixed_indices,
                        alpha_tilde = NULL,
                        Z_tilde_list = NULL,
                        barrier_t = 1, #starting value of reciprocal barrier penalty coef.
                        barrier_scale = 10, #increments for value of barrier penalty
                        max_barrier = 1e12, #maximum value of barrier_t
                        initial_conv_tol = 1000,
                        final_conv_tol = 0.1,
                        constraint_tolerance = 1e-10,
                        hessian_regularization = 0.01,
                        criterion = "Poisson",
                        profile_P = FALSE,
                        profiling_maxit = 25,
                        wts = NULL,
                        verbose = FALSE)
  
}

full_cv_predictions <- lapply(1:10,
                              function(x)
                                full_cv[[x]]$varying[
                                  full_cv[[x]]$varying$param == "P"&
                                    full_cv[[x]]$varying$k>1,])

for(i in 1:10){
  full_cv_predictions[[i]]$k <- sapply(full_cv_predictions[[i]]$k,
                                       function(x) folds[[i]][x - 1])
  
}

full_cv_predictions <- do.call(rbind,full_cv_predictions)
fc_values <- W[1,]/sum(W[1,])
full_cv_predictions$fc_value <- sapply(full_cv_predictions$j,
                                       function(d) fc_values[d])
full_cv_predictions$protocol <-
  sapply(full_cv_predictions$k,
         function(d) protocols[d-1]) #  d - 1 bc k starts at 2
#  (k = 1 is fc data)


full_cv_predictions$specimen <-
  sapply(full_cv_predictions$k,
         function(d) individuals[d - 1]) #  d - 1 bc k starts at 2
#  (k = 1 is fc data)


null_cv_predictions <- lapply(1:10,
                              function(x)
                                null_cv[[x]]$varying[
                                  null_cv[[x]]$varying$param == "P"&
                                    null_cv[[x]]$varying$k>1,])

for(i in 1:10){
  null_cv_predictions[[i]]$k <- sapply(null_cv_predictions[[i]]$k,
                                       function(x) folds[[i]][x - 1])
  
}

null_cv_predictions <- do.call(rbind,null_cv_predictions)
fc_values <- W[1,]/sum(W[1,])
null_cv_predictions$fc_value <- sapply(null_cv_predictions$j,
                                       function(d) fc_values[d])
null_cv_predictions$protocol <-
  sapply(null_cv_predictions$k,
         function(d) protocols[d -1 ]) #  d - 1 bc k starts at 2 (k = 1 is fc data)

null_cv_predictions$specimen <-
  sapply(null_cv_predictions$k,
         function(d) individuals[d - 1]) #  d - 1 bc k starts at 2 (k = 1 is fc data)
null_cv_predictions$model <- "Null Model"
full_cv_predictions$model <- "Full Model"

W_prop <- W[-1,]
for(i in 1:nrow(W_prop)){
  W_prop[i,] <- W_prop[i,]/sum(W_prop[i,])
}

naive_predictions <- null_cv_predictions[numeric(0),]

for(i in 1:nrow(W_prop)){
  protocol <- protocols[i]
  specimen <- individuals[i]
  for(j in 1:ncol(W)){
    naive_predictions <- rbind(naive_predictions,
                               data.frame(value = W_prop[i,j],
                                          param = "P",
                                          k = i,
                                          j  = j,
                                          fc_value = fc_values[j],
                                          protocol = protocol,
                                          specimen = specimen,
                                          model = "Plug-in"))
  }
}




rbind(null_cv_predictions,full_cv_predictions,
      naive_predictions) %>%
  mutate(protocol = sapply(protocol, function(x) paste("Protocol ",x,
                                                       sep = "",
                                                       collapse = ""))) %>%
  # filter(!is.na(protocol)) %>% #why would protocol be NA? Check!
  ggplot() +
  geom_point(aes(x = fc_value, y = value, color= specimen),
             size = .5) +
  geom_line(aes(x = fc_value, y = value, color = specimen,
                group = as.factor(k)), size = .5) +
  geom_abline(aes(intercept = 0, slope = 1), linetype = 2) +
  facet_grid(model~protocol,scales = "free_y") +
  scale_color_grey() +
  scale_y_log10() +
  scale_x_log10() +
  theme_bw() +
  guides(color=guide_legend(title="Specimen")) +
  xlab("Relative Abundance Measured by Flow Cytometry") +
  ylab("Cross-Validated Estimated Relative Abundance")
```

## References

- Costea, P., Zeller, G., Sunagawa, S. et al. Towards standards for human fecal sample processing in metagenomic studies. *Nat Biotechnol* **35**, 1069–1076 (2017). https://doi.org/10.1038/nbt.3960. 